Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-07-04 interimreport1.Rnw
Interim report

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{bibunits}
%\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}

@@

\title{Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\date{\today}  %\today is replaced with the current date
\maketitle

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modelling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more. Our key findings were:
\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one. This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
\end{itemize}

@@

The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997. Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions. The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{\today}). In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.

Previous analyses of measles, including two in New Zealand by Prof. Roberts \cite{roberts0,roberts4}, have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.

We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our preliminary modelling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modelling}).

@@

\subsection{Additional risk analyses}\label{sec:risk_analyses}

@@

For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note this uses all immigration of foreign nationals, coming for whatever purpose. We used the WHO data to determine per capita measles cases for each year and used this and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.

@@

\begin{table}
\caption{Lowest national measles vaccine cover (2012)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
@
\label{table:cover12}
\end{table}

@@

Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}. However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:incidence12} and Table \ref{table:cover12}). Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table \ref{table:risk12}. The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.

@@

\subsection{Future risk analyses}\label{sec:risk_analyses}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above. These data only became available to us on 3 July 2014, provided by the University of Otago.

@@

Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \cite{clements4}. Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities. \cite{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005. Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005. They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination. The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \cite{who13}.  In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.

The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million \cite{carabin3]. The estimated cost for a case ranged from US\$ 189--344 \cite{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$ 967--1,755 \cite{carabin2}. \cite{stack11} estimated the economic benefits from cases averted due to measles vaccination. They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020. Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.

Italy has the highest reported annual cost of measles among industrialised countries \cite{carabin3}. In 2001, it reported losses related to measles of approximately US\$ 50 million. The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis. The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days). The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$ 2.0 in 2002-03), or approximately \euro 1,700 per case. The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721. The average daily cost of a hospital stay was  \euro 327.

An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \cite{flego13}. Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department. During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney. The mean number of contacts per case was 54 (median = 28, maximum = 206). The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$ 48,000, with 90\% of this being associated with staff time.

Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years. In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \cite{wichmann9}. However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006). An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW). In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case. Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised. The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623. In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing. Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case. In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.

Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \cite{parker6}. Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.  \cite{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011. The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$ 2.7 and 5.3 million. Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$ 298 per contact.
\cite{dayan5} calculated the cost of containing a single case of measles that occurred in Iowa in 2004. They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination. In total, it was estimated the direct costs associated with three cases of measles was US\$ 142,452, or nearly US\$ 50,000 per case.

\cite{parker6} reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana. A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak. Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak. Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead. This amounted to a direct cost of \$4,932 per measles case. These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.

\cite{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.  Costs included labor, translation and benefits for public health workers. In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis. In total, 387 hours were associated with this single case, resulting in a cost of US\$ 11,881. In addition, per-contact costs amounted to US\$ 264. The cost of hospitalisation for the 3-day stay by the index case was US\$ 931. Additional costs were associated with physician visits (US\$ 294), vaccine and immunoglobulin (US\$ 1,765), mileage (US\$ 205) and immunologic screening tests for the parents' exposed to measles (US\$ 240) for a total of US\$ 23,816.

@@

\subsection{Methods}
Costs were evaluated as either direct or indirect. Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs. Direct costs can be divided into medical and non-medical \cite{saha13}. Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care. Personnel time (investigation and emergency response), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs. Direct non-medical costs include transportation to and from health care providers.

@@

In other outbreaks, the average cost per measles case was estimated to be US\$ 254, US\$ 276, and US\$ 307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002). This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$ 142,542. In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005). They estimated direct costs per case to be less than US\$ 500. 
The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty1}. In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person. In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864. The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days). They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \cite{zhou4}.

\subsection{Cost analysis summary}
\begin{itemize}
\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$ 750,000.
\end{itemize}

\subsection{Future cost-benefit analyses}
Using the results above we aim to:
\begin{itemize}
\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
\end{itemize}

We also require additional clarifications of the data, regarding what hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours? Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.

\section{Modelling measles epidemics}\label{sec:epidemic_modelling}

A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \cite{roberts0,roberts4,tobias98}. The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.

The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \cite{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \cite{mansoor98,roberts0}. Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics. The model developed by \cite{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five. The schedule was changed in 2000 with MMR2 now being administered before 5 years \cite{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics.

>>>>

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
%\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{eurosym}
\usepackage{todonotes}

@@

\title{Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\date{3 July 2014}  %\today is replaced with the current date
\maketitle

@@

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad. In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modelling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more. Our key findings were:
\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one. This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
\end{itemize}

@@

The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997. Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions. The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}). In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.

We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe. Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control. Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007). However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving. This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).

@@

\subsection{Additional risk analyses}
\label{sub:risk_analyses}

@@

For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand. We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Note this uses all immigration of foreign nationals, coming for whatever purpose. We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.

@@

\begin{table}
\caption{Lowest national measles vaccine cover (\%, 2012)}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
@
\label{table:cover12}
\end{table}

@@

Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}. However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}). Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table \ref{table:risk12}. The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.

@@

\subsection{Future risk analyses}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data only became available to us on 3 July 2014, provided by the University of Otago.

@@

Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \citep{clements4}. Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities. \citep{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005. Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005. They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination. The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.  In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.

The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3]. The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}. \citep{stack11} estimated the economic benefits from cases averted due to measles vaccination. They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020. Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.

Italy has the highest reported annual cost of measles among industrialised countries \citep{carabin3}. In 2001, it reported losses related to measles of approximately US\$50 million. The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis. The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days). The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$2.0 in 2002-03), or approximately \euro 1,700 per case. The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721. The average daily cost of a hospital stay was  \euro 327.

An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \citep{flego13}. Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department. During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney. The mean number of contacts per case was 54 (median = 28, maximum = 206). The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$48,000, with 90\% of this being associated with staff time.

Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years. In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \citep{wichmann9}. However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children \citep{siedler6}. An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW). In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case. Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised. The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623. In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing. Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case. In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.

Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \citep{parker6}. Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.  \citep{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011. The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$2.7 and 5.3 million. Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298 per contact.
\citep{dayan5} calculated the cost of containing a single case of measles that occurred in Iowa in 2004. They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination. In total, it was estimated the direct costs associated with three cases of measles was US\$142,452, or nearly US\$50,000 per case.

\citep{parker6} reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana. A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak. Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak. Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead. This amounted to a direct cost of \$4,932 per measles case. These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.

\citep{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.  Costs included labor, translation and benefits for public health workers. In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis. In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881. In addition, per-contact costs amounted to US\$264. The cost of hospitalisation for the 3-day stay by the index case was US\$931. Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents' exposed to measles (US\$240) for a total of US\$23,816.

@@

\subsection{Methods}
Costs were evaluated as either direct or indirect. Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs. Direct costs can be divided into medical and non-medical \citep{saha13}. Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care. Personnel time (investigation and emergency response), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs. Direct non-medical costs include transportation to and from health care providers.

@@

In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002). This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542. In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005). They estimated direct costs per case to be less than US\$500. 
The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}. In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person. In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864. The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days). They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.

\subsection{Cost analysis summary}
\begin{itemize}
\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$750,000.
\end{itemize}

\subsection{Future cost-benefit analyses}
Using the results above we aim to:
\begin{itemize}
\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
\end{itemize}

We also require additional clarifications of the data, regarding:
\begin {itemize}
\item What hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours?
\end {itemize}
Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.

\section{Modelling measles epidemics}
\label{sec:epidemic_modelling}

A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}. The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.

The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}. Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics. The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five. The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.

These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada. In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel. All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics. \citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one. They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
